---
title: "Final Project Report on Practical Machine Learning Course"
author: "Ritika Pandey"
date: "May 1, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Introduction 

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).

#Data Sources

The training data for this project are available here:

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

The test data are available here:

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

The data for this project come from this source: http://groupware.les.inf.puc-rio.br/har. We are very thankful to them as they have been very generous in allowing their data to be used for this kind of assignment.

#Project Goal

The goal of this project is to predict the manner in which the 6 participants did the exercise. This is the "classe" variable in the training set. You may use any of the other variables to predict with. Create a report describing how model is built, usage cross validation, what is expected out of sample error, and justify the choices made. Use prediction model to predict 20 different test cases.

#Preliminary Work

###_Reproduceabliity_
An overall pseudo-random number generator seed was set at 1234 for all code. In order to reproduce the results below, the same seed should be used. Different packages were downloaded and installed, such as caret and randomForest. These should also be installed in order to reproduce the results below (please see code below for ways and syntax to do so)

###_Approach_
Our outcome variable is classe, a factor variable. For this data set, "participants were asked to perform one set of 10 repetitions of the Unilateral Dumbbell Biceps Curl in 5 different fashions: - exactly according to the specification (Class A) - throwing the elbows to the front (Class B) - lifting the dumbbell only halfway (Class C) - lowering the dumbbell only halfway (Class D) - throwing the hips to the front (Class E).
Two models will be tested using decision tree and random forest algorithms. The model with the highest accuracy will be chosen as our final model.

###_Cross Validation_
Cross-validation will be performed by subsampling our training data set randomly without replacement into 2 subsamples: subTraining data (75% of the original Training data set) and subTesting data (25%). Our models will be fitted on the subTraining data set, and tested on the subTesting data. Once the most accurate model is choosen, it will be tested on the original Testing data set.

###_Expected out-of-sample error_
The expected out-of-sample error will correspond to the quantity: 1-accuracy in the cross-validation data.Accuracy is the proportion of correct classified observation over the total sample in the subTesting data set.Expected accuracy is the expected accuracy in the out-of-sample data set (i.e. original testing data set). Thus, the expected value of the out-of-sample error will correspond to the expected number of missclassified observations/total observations in the Test data set, which is the quantity: 1-accuracy found from the cross-validation data set.

Our outcome variable "classe" is an unordered factor variable.Thus, we can choose our error type as 1-accuracy.We have a large sample size with N= 19622 in the Training data set.This allow us to divide our Training sample into subTraining and subTesting to allow cross-validation.Predictors with all missing values will be discared as well as predictors with irrelevant values.Decision tree and random forest algorithms are known for their ability of detecting the features that are important for classification.

#Packges,set.seed and libraries 

Installing packages, loading libraries, and setting the seed for **reproduceability** : 

```{r}
library(caret)
library(ggplot2)
library(randomForest)
library(rpart)
library(rpart.plot)
library(rattle)
library(RColorBrewer)
set.seed(12345)
```

#Getting and Cleaning Data

The training data set can be found on the following URL:

```{r}
trainUrl <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
```
The testing data set can be found on the following URL:
```{r}
testUrl <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
```
Load data to memory.

```{r}
training <- read.csv(url(trainUrl), na.strings=c("NA","#DIV/0!",""))
testing <- read.csv(url(testUrl), na.strings=c("NA","#DIV/0!",""))
```
Delete columns with all missing values
```{r}
training<-training[,colSums(is.na(training))==0]
testing<-testing[,colSums(is.na(testing))==0]
```
Delete variables that are irrelevant to our project: user_name, raw_timestamp_part_1, raw_timestamp_part_,2 cvtd_timestamp, new_window, and  num_window (columns 1 to 7)

```{r}
training   <-training[,-c(1:7)]
testing <-testing[,-c(1:7)]
```
#Partitioning training dataset

Partioning Training data set into two data sets, 75% for subTraining, 25%% for subTesting:
```{r}
inTrain <- createDataPartition(y=training$classe, p=0.75, list=FALSE)
subTraining<-training[inTrain,]
subTesting<-training[-inTrain,]
```

```{r}
dim(subTraining)
dim(subTesting)

##The variable "classe" contains 5 levels: A, B, C, D and E. A plot of the outcome variable will allow us to see the frequency of each levels in the subTraining data set and # compare one another.

plot(subTraining$classe,col="green",main="Plot of levels of variable classe",xlab="classe",ylab="frequency")
```
Based on the graph above, we can see that each level frequency is within the same order of magnitude of each other. Level A is the most frequent while level D is the least frequent.

#Prediction Model 1 : Decision Tree

```{r}
model1<-rpart(classe~.,data=subTraining,method = "class")

#To view the decision tree
fancyRpartPlot(model1)
```

Predicting: 

```{r}
prediction1<-predict(model1,subTesting,type = "class")
```
Test result using ConfusionMatrix:
```{r}
confusionMatrix(prediction1, subTesting$classe)
```
#Prediction Model 2: Random Forest 

```{r}
model2<-randomForest(classe~.,data=subTesting,method="class")
```

Predicting;

```{r}
prediction2<-predict(model2,subTesting,type="class")
```
Test result using ConfusionMatrix:
```{r}
confusionMatrix(prediction2,subTesting$classe)
```

#Which Prediction Model to use?

Random Forest algorithm performed better than Decision Trees. Accuracy for Random Forest model was 1 (95% CI: (0.9992, 1)) compared to Decision Tree model with  0.6407 (95% CI: (0.6271, 0.6541)). The Random Forests model is choosen. The expected out-of-sample error is estimated at 0.005, or 0.5%.

#Submission 

Final outcome based on Prediction Model 2 i.e. Random Forest applied on Testing Dataset : 

```{r}
finalPrediction<-predict(model2,testing,type="class")
finalPrediction
```

